Using tours to visually investigate properties of new projection pursuit indexes with application to problems in physics

### 2019-01-17 scagnostics.Rmd
done with conclusions

Physics examples motivate us to investigate projection pursuit indexes that detect complex patterns in two dimensional projections. In this context we have investigated the use of indexes developed for variable selection as projection pursuit indexes. The set of relevant criteria that define a good index function includes: smoothness, squintability, flexibility, rotation invariance and speed. To evaluate these criteria we used three different sets of simulated data, defined to have different underlying noise distributions (uniform or normal) and different structures (circle, sine and spiral which can be considered as large, medium and fine structure respectively). The properties were assessed by a list of procedures introduced in Section \ref{sec:propAss} and we have evaluated several index functions proposed in the literature for variable selection: scagnostics indexes, maximum information based indexes, distance correlation and a spline based index. While all index functions are found to fall short on at least some of the criteria, we have shown that combined with the optimisation available via the guided tour they can already be used to identify interesting views, in particular the built-in structures in the simulated data sets.

@@

As a final remark we point out that small changes to the implementation of the guided tour algorithm would already be helpful, and we plan on implementing the following: allowing for a stopping condition based on index value, and making the step size for the selection of the most promising direction an argument when calling the geodesic optimiser.

>>>>

<!--
Reminder what the paper is about
summarise main findings
Future plans
-->

The motivation for this work was to discover dependencies between estimated parameters in multiple model fits in physics problems. This paper shows how projection pursuit with the new indexes can help address this problem. The results are encouraging, showing large potential for discovering unanticipated relations between multiple variables.

All of the indexes fall short against some aspect of the ideal properties of smoothness, squintability, flexibility, rotation invariance and speed. The paper describes how these properties can be assessed using tour methodology. Some potential fixes for the indexes are discussed but there is scope for further developing the new indexes.

The work also reveals inadequacies in the tour optimization algorithm, that may benefit from newly developed techniques and software tools. Exploring this area would help improve the guided tours. The tour is also lacking a portable, easy to use interface that would enable user interactions like stopping, re-starting, saving optimal projections, and development of a graphical user interface, similar to the tourrGui [@tourrGui] (now archived), would help users and projection pursuit researchers.

<!--
Physics examples motivate us to investigate projection pursuit indexes that detect complex patterns in two dimensional projections. In this context we have investigated the use of indexes developed for variable selection as projection pursuit indexes. The set of relevant criteria that define a good index function includes: smoothness, squintability, flexibility, rotation invariance and speed. To evaluate these criteria we used three different sets of simulated data, defined to have different underlying noise distributions (uniform or normal) and different structures (circle, sine and spiral which can be considered as large, medium and fine structure respectively). The properties were assessed by a list of procedures introduced in Section \ref{sec:propAss} and we have evaluated several index functions proposed in the literature for variable selection: scagnostics indexes, maximum information based indexes, distance correlation and a spline based index. While all index functions are found to fall short on at least some of the criteria, we have shown that combined with the optimisation available via the guided tour they can already be used to identify interesting views, in particular the built-in structures in the simulated data sets.

@@

As a final remark we point out that small changes to the implementation of the guided tour algorithm would already be helpful, and we plan on implementing the following: allowing for a stopping condition based on index value, and making the step size for the selection of the most promising direction an argument when calling the geodesic optimiser.
-->
